---
author: Nasir Abdulrasheed
title: Seasonality Detection Pipeline - Test
execute: 
  echo: false
  warning: false
  message: false
format:
  html: 
    toc: true
    colorlinks: true
    number-sections: false
    fig-pos: H
    code-fold: true
    smooth-scroll: true
    respect-user-color-scheme: true
    theme:
     light: litera
     dark: cyborg
---

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(arrow)
library(gt)
library(webshot2)
library(fpp3)
library(bizdays)
library(timeDate)
library(patchwork)
library(here)
library(glue)
library(purrr)
library(urca)
library(ggrepel)

FLUPENTIXOL <- "0402020G0%"
PALIPERIDONE <- "0402020AB%"
ZUCLOPENTHIXOL <- "0402020Z0%"
NORETHISTERONE <- "0604012P0%"
CELECOXIB <- "1001010AH%"

PLOTS_DIR <- "outputs/nasir/test/figures"
TBL_DIR <- "outputs/nasir/test/tables"

DETREND_TYPE <- "stl"

DATE_START <- "2016-05-01"
DATE_END <- "2026-04-01"
```

# Data

```{r}
#| label: tbl-load-data
#| tbl-cap: Monthly prescribing data for select drugs (May 2016 to April 2026)

test_drugs_file <- glue("df_test_drugs_{DATE_START}_{DATE_END}.parquet")
DF_PRESC <- read_parquet(here("data-raw", "nasir", test_drugs_file))

N_MONTHS <- unique(DF_PRESC$month) |> length()
N_YEAR_MONTHS <- 12 * (N_MONTHS %/% 12)
```

The following medications (depot antipsychotics, NSAIDs, and Contraceptives) were used to test the pipline and the results displayed here:

1. Flupentixol (0402020G0)
2. Paliperidone (0402020AB)
3. Zuclopenthixol (0402020Z0)
4. Norethisterone (0604012P0)
5. Celecoxib (1001010AH)

The data used spans exactly `r {N_MONTHS}` months (`r ({N_MONTHS / 12})` years).

```{r}
#| label: func-work-days
#| warning: false

# bank holidays in the data period
uk_holidays <- as.Date(holidayLONDON(year(DATE_START):year(DATE_END)))

dispensing_days <- create.calendar(
  name = "UK",
  weekdays = c("sunday"),
  holidays = uk_holidays
)

month_work_days <- function(date) {
  month_start <- as.Date(date)
  month_end <- seq(month_start, by = "month", length.out = 2)[2] - 1
  length(bizseq(month_start, month_end, dispensing_days))
}
```

```{r}
#| label: func-aggregate-data

aggregate_ts <- function(df) {
  df |>
    summarise(
      total_items = sum(items),
      .by = c(month, chemical_code, chemical)
    ) |>
    mutate(month = yearmonth(month)) |>
    as_tsibble(index = month, key = c(chemical, chemical_code)) |>
    mutate(items_per_wd = total_items / map_dbl(month, month_work_days))
}
```

```{r}
#| label: func-visualise

visualise_ts <- function(ts) {
  if (!inherits(ts, "tbl_ts")) {
    stop("Argument `ts` must be a time series object")
  }

  chems <- unique(ts$chemical)
  plots <- list()
  for (i in seq_along(chems)) {
    chem <- chems[i]
    chem_ts <- filter(ts, chemical == chem)
    time_plot <- chem_ts |>
      autoplot(items_per_wd) +
      theme(legend.position = "none") +
      theme_minimal() +
      labs(y = "Items per work day", x = "Month [1M]")

    seasonal_plot <- chem_ts |>
      gg_season(
        items_per_wd,
        labels = "right",
        labels_repel = TRUE
      ) +
      theme_minimal() +
      labs(y = "Items per work day", x = "Month [1M]")

    subseries_plot <- chem_ts |>
      gg_subseries(items_per_wd) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text.y = element_blank(),
        strip.background = element_blank()
      ) +
      labs(y = "Items per work day", x = "Year")

    acf_plot <- chem_ts |>
      ACF(items_per_wd) |>
      autoplot() +
      theme_minimal() +
      labs(y = "ACF", x = "Lag [1M]")

    p <- (time_plot / seasonal_plot / subseries_plot) +
      plot_layout(axis_titles = "collect_y")
    p <- (p / acf_plot) +
      plot_annotation(title = chem)

    save_file <- glue("{chem}_plots_{DATE_START}_{DATE_END}.png")
    ggsave(
      here(PLOTS_DIR, save_file),
      width = 7,
      height = 7,
      units = "in"
    )

    plots[[i]] <- p
  }

  plots
}
```

```{r}
#| label: func-detrend

detrend_stl <- function(ts, s.window = 13) {
  DETREND_TYPE <<- "stl"
  ts |>
    model(
      STL(items_per_wd ~ trend() + season(window = s.window))
    ) |>
    components() |>
    mutate(detrended = season_year + remainder) |>
    select(chemical, chemical_code, month, detrended, season_year)
}

detrend_linear <- function(ts) {
  # simple linear time trend
  DETREND_TYPE <<- "linear"

  ts |>
    model(linear_trend = TSLM(items_per_wd ~ trend())) |>
    residuals() |>
    select(chemical, chemical_code, month, detrended = .resid)
}

detrend_loess <- function(ts, span = 0.75) {
  # plain loess detrending
  DETREND_TYPE <<- "loess"

  ts |>
    as_tibble() |>
    group_by(chemical, chemical_code) |>
    mutate(
      t = row_number(),
      # Fit the Loess curve and extract the fitted trend line
      trend_curve = loess(items_per_wd ~ t, span = span)$fitted,
      detrended = items_per_wd - trend_curve
    ) |>
    ungroup() |>
    select(-t, -trend_curve) |>
    as_tsibble(index = month, key = c(chemical, chemical_code))
}

plot_decomp <- function(ts_decomp) {
  chems <- unique(ts$chemical)
  for (chem in chems) {
    chem_decomp <- ts_decomp |>
      filter(chemical == chem)

    (chem_decomp |>
      autoplot(items_per_wd) +
      labs(title = chem)) |>
      print()
  }
}

detrended_plots <- function(ts_detrend) {
  rows <- length(unique(ts_detrend$chemical))
  p <- ggplot(ts_detrend) +
    geom_line(aes(x = month, y = detrended)) +
    facet_wrap(~chemical, scales = "free", nrow = rows) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(
      x = "Month",
      y = "Items per wd"
    )

  ggsave(
    here(PLOTS_DIR, glue("detrended_plots_{DETREND_TYPE}.png"))
  )

  p
}

seasonal_comp_plots <- function(ts_detrend) {
  rows <- length(unique(ts_detrend$chemical))
  p <- ggplot(ts_detrend) +
    geom_line(aes(x = month, y = season_year)) +
    facet_wrap(~chemical, scales = "free", nrow = rows) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(
      x = "Month",
      y = "Items per wd"
    )

  ggsave(
    here(PLOTS_DIR, glue("seasonal_comp_plots_{DETREND_TYPE}.png"))
  )

  p
}

strength_stl <- function(ts_agg) {
  FIRST_MONTH <- ts_agg |>
    slice_head(n = 1) |>
    pull(month) |>
    month()

  results <- ts_agg |>
    features(items_per_wd, feat_stl) |>
    select(-spikiness, -linearity, -curvature, -starts_with("stl_")) |>
    mutate(
      seasonal_peak = month.abb[
        ((FIRST_MONTH - 1 + seasonal_peak_year) %% 12) + 1
      ],
      seasonal_trough = month.abb[
        ((FIRST_MONTH - 1 + seasonal_trough_year) %% 12) + 1
      ]
    ) |>
    select(-seasonal_peak_year, -seasonal_trough_year)

  write_parquet(
    results,
    here(TBL_DIR, "strength_scores.parquet")
  )

  results
}
```

```{r}
#| label: fft-functions

compute_fft <- function(ts) {
  if (!is.vector(ts)) {
    stop("Argument `ts` must be a vector")
  }

  len_ts <- length(ts)

  ts <- ts[(len_ts - N_YEAR_MONTHS):len_ts]
  len_ts <- N_YEAR_MONTHS / 2

  amps <- fft(ts) |> abs()

  amps[2:(len_ts + 1)]
}

filter_fft <- function(fft_amps, top_k = 1) {
  if (!is.vector(fft_amps)) {
    stop("Argument `fft_amps` must be a vector")
  }

  if (top_k > length(fft_amps)) {
    stop("`top_k` cannot exceed the length of `fft_amps` amplitudes vector")
  }

  sort_idx <- order(fft_amps, decreasing = TRUE)
  min_freq <- N_YEAR_MONTHS %/% 24
  max_freq <- N_YEAR_MONTHS %/% 2
  sort_idx <- sort_idx[(sort_idx >= min_freq) & (sort_idx <= max_freq)]
  candidates <- sort_idx[1:top_k]
  amps <- fft_amps[candidates]

  list(freqs = candidates, amps = amps)
}

plot_fft <- function(fft_amps, chemical) {
  if (!is.vector(fft_amps)) {
    stop("Argument `fft_amps` must be a vector")
  }

  len_ts <- length(fft_amps)

  freqs <- 1:len_ts

  spectral_plot <- ggplot(mapping = aes(x = freqs, y = fft_amps)) +
    geom_segment(
      aes(xend = freqs, y = 0, yend = fft_amps),
      colour = "steelblue"
    ) +
    geom_point(colour = "steelblue", size = 1) +
    theme_minimal() +
    labs(
      x = "Frequency",
      y = "Power",
      title = chemical
    )

  save_file <- glue("{chemical}_fft_{DETREND_TYPE}_{DATE_START}_{DATE_END}.png")
  ggsave(
    here(PLOTS_DIR, save_file),
    width = 7,
    height = 7,
    units = "in"
  )

  spectral_plot
}
```

```{r}
#| label: func-fft-score

score <- function(ts_dable, top_k = 1) {
  chems <- unique(ts_dable$chemical)
  scores <- tibble()
  plots <- list()

  for (i in seq_along(chems)) {
    chem <- chems[i]
    chem_dable <- ts_dable |>
      filter(chemical == chem)

    fft_amps <- chem_dable |>
      pull(detrended) |>
      compute_fft()

    plots[[i]] <- plot_fft(fft_amps, chem)

    amps_freqs <- filter_fft(fft_amps, top_k)
    tmp_df <- tibble(chemical = chem)

    for (j in seq_along(amps_freqs$freqs)) {
      col_period <- paste0("period_", j)
      col_freq <- paste0("freq_", j)
      col_score <- paste0("score_", j)

      tmp_df <- bind_cols(
        tmp_df,
        tibble(
          "{col_period}" := round(N_YEAR_MONTHS / amps_freqs$freqs[j], 2),
          "{col_freq}" := round(amps_freqs$freqs[j], 2),
          "{col_score}" := amps_freqs$amps[j]
        )
      )
    }

    scores <- bind_rows(scores, tmp_df)
  }

  list(plots = plots, scores = scores)
}
```

```{r}
#| label: func-modelling
#| warning: false
#| error: false

fit_models <- function(ts, df_periods, max_k = 6) {
  period_grid <- df_periods |>
    pivot_longer(
      cols = starts_with("period_"),
      names_to = "period_name",
      values_to = "period",
      values_drop_na = TRUE
    ) |>
    mutate(max_allowed_k = pmin(max_k, period %/% 2)) |>
    select(chemical, period, max_allowed_k)

  map_dfr(unique(df_periods$chemical), function(chem) {
    ts_chem <- ts |> filter(chemical == chem)

    # compute model K = 0 once, not dependent on P
    m0 <- ts_chem |> model(ARIMA(items_per_wd ~ PDQ(0, 0, 0)))
    g0 <- glance(m0) |>
      mutate(
        .model = "K0",
        period = NA,
        k = 0,
        lrt_stat = NA,
        p_value = NA,
        .after = .model
      )

    periods_chem <- period_grid |>
      filter(chemical == chem) |>
      select(period, max_allowed_k)

    models_df <- pmap_dfr(periods_chem, function(period, max_allowed_k) {
      if (max_allowed_k < 1) {
        return(tibble())
      }

      prev_log_lik <- g0$log_lik
      prev_bic <- g0$BIC
      rows <- vector("list", max_allowed_k)

      for (k in 1:max_allowed_k) {
        form <- items_per_wd ~ fourier(period = period, K = k) + PDQ(0, 0, 0)
        m <- ts_chem |> model(ARIMA(form))

        g <- glance(m)
        g <- mutate(
          .data = g,
          .model = glue("P{period}_K{k}"),
          period = period,
          k = k,
          # D statistic: -2 * (L0 - L1)
          lrt_stat = -2 * (prev_log_lik - g$log_lik),
          p_value = pchisq(lrt_stat, df = 2, lower.tail = FALSE),
          .after = .model
        )

        # Short-circuit: stop growing this period's harmonic ladder the
        # moment another Fourier pair fails to improve BIC
        if (g$BIC >= prev_bic) {
          break
        }

        rows[[k]] <- g

        prev_log_lik <- g$log_lik
        prev_bic <- g$BIC
      }

      bind_rows(rows)
    })

    bind_rows(g0, models_df) |>
      mutate(chemical = chem, .before = 1) |>
      select(-ar_roots, -ma_roots) |>
      filter(is.na(p_value) | p_value < 0.05) |>
      mutate(
        p_value = if_else(
          !is.na(p_value),
          format.pval(p_value, digits = 2),
          "-"
        )
      ) |>
      group_by(chemical_code, chemical) |>
      arrange(BIC)
  })
}
```

```{r}
#| label: func-report

generate_report <- function(
  ts_original,
  candidates,
  models,
  strength,
  top_k = 5
) {
  period_cols <- paste0("period_", 1:top_k)
  candidates_long <- candidates |>
    pivot_longer(
      cols = all_of(period_cols),
      names_to = "rank",
      values_to = "period",
      values_drop_na = TRUE
    ) |>
    mutate(rank = as.integer(str_extract(rank, "\\d+"))) |>
    arrange(chemical, rank)

  # FFT dominant periods
  dom_period <- candidates_long |>
    mutate(period = round(period, 1)) |>
    summarise(
      dominant_periods = paste(head(period, 3), collapse = ", "),
      .by = "chemical"
    )

  best_models <- models |>
    group_by(chemical) |>
    slice_head(n = 1) |>
    ungroup() |>
    select(chemical, .model, period, k, p_value)

  # component periods contained in the winning model:
  # period/1, period/2, ..., period/k.
  model_periods <- best_models |>
    mutate(harmonic = map(k, seq_len)) |>
    unnest(harmonic) |>
    mutate(component_period = round(period / harmonic, 1)) |>
    arrange(chemical, harmonic) |>
    summarise(
      model_period_list = paste(component_period, collapse = ", "),
      n_components = n(),
      .by = "chemical"
    )

  strength_info <- strength |>
    select(chemical, seasonal_strength_year, seasonal_peak, seasonal_trough)

  report_tbl <- dom_period |>
    left_join(best_models, by = "chemical") |>
    left_join(model_periods, by = "chemical") |>
    left_join(strength_info, by = "chemical") |>
    mutate(
      model_info = glue("{.model} (P={period}, K={k})"),
      .keep = "unused"
    ) |>
    select(
      Chemical = chemical,
      `Dominant periods (FFT)` = dominant_periods,
      `Model periods` = model_period_list,
      `Peak month` = seasonal_peak,
      `Trough month` = seasonal_trough,
      `Seasonal strength` = seasonal_strength_year,
      `Best model` = model_info,
      `LRT p-value` = p_value
    )

  report_tbl
}
```

# Full Pipeline
Putting it all together.

### Aggregation and Visualisation 

```{r}
#| label: tbl-aggregated-data
#| tbl-cap: Sample of aggregated time series data

DF_PRESC_AGG <- aggregate_ts(DF_PRESC)

head(DF_PRESC_AGG, 5) |> gt()
```

```{r}
#| label: fig-vis
#| fig-cap: Plots of original time series

purrr::walk(visualise_ts(DF_PRESC_AGG), print)
```

```{r}
#| label: data-length-check
#| output: asis

if (N_YEAR_MONTHS < 60) {
  cat("::: {.callout-important}\n")
  cat("## Halted\n\n")
  cat("**Analysis terminated:** fewer than 5 years of data are available.\n")
  cat(":::\n")

  knitr::knit_exit()
}
```

### Detrending

Strengh scores
```{r}
#| label: tbl-strength-scores
#| tbl-cap: Seasonal and trend strength using `feasts::feat_stl()`

strength_scores <- strength_stl(DF_PRESC_AGG)

gt(strength_scores)
```

```{r}
DF_PRESC_AGG <- DF_PRESC_AGG |>
  left_join(strength_scores, by = c("chemical_code", "chemical")) |>
  filter(seasonal_strength_year >= 0.33)
```

```{r}
#| label: fig-detrended-plots
#| fig-cap: Detrended time plots

df_detrended <- detrend_stl(DF_PRESC_AGG)

detrended_plots(df_detrended)
```

```{r}
#| label: fig-seasonal-comp-plots
#| fig-cap: Plots of the pure seasonal components

seasonal_comp_plots(df_detrended)
```

### FFT
```{r}
#| label: fig-spectral-plots
#| fig-cap: Spectral plots

TOP_K <- 5
fft_scores <- score(df_detrended, top_k = TOP_K)

wrap_plots(fft_scores$plots, ncol = 2)
```

### Candidate periods
```{r}
#| label: tbl-fft-periods
#| tbl-cap: !expr glue("Candidate periods found and their signal strength from FFT (Top {TOP_K})")

candidates <- fft_scores$scores

write_parquet(
  candidates,
  here(TBL_DIR, glue("candidate_periods_{DETREND_TYPE}.parquet"))
)
gt(candidates)
```

### Modelling

```{r}
#| label: tbl-modelling
#| tbl-cap: Top 3 performing models for each chemical
#| #| warning: false
#| error: false

models_results <- DF_PRESC_AGG |>
  fit_models(candidates)

write_parquet(
  models_results,
  here(TBL_DIR, glue("modelling_results_{DETREND_TYPE}.parquet"))
)

slice_head(models_results, n = 3) |> gt()
```

### Summary report

```{r}
#| label: report

summary_tbl <- generate_report(
  ts_original = DF_PRESC_AGG,
  candidates = candidates,
  models = models_results,
  strength = strength_scores,
  top_k = TOP_K
)

write_parquet(
  summary_tbl,
  here(TBL_DIR, glue("summary_table_{DETREND_TYPE}.parquet"))
)

gt(summary_tbl) |>
  tab_header(
    title = "Seasonality Detection Summary",
    subtitle = glue(
      "Data period: {DATE_START} to {DATE_END} | Detrending: {DETREND_TYPE}"
    )
  ) |>
  fmt_number(columns = `Seasonal strength`, decimals = 2) |>
  tab_style(
    style = cell_fill(color = "#e6f2ff"),
    locations = cells_body(columns = `Model periods`)
  ) |>
  tab_footnote(
    footnote = paste0(
      "Dominant periods from FFT (strongest amplitudes), shown independently as a sanity check. ",
      "Model periods are the harmonic components (period/1, period/2, ...) actually retained ",
      "in the BIC-selected model. Peak/Trough months are the calendar months with highest/lowest ",
      "average dispensing. Best model from ARIMA errors + Fourier terms, selected by BIC. ",
      "Seasonal strength from STL decomposition."
    )
  ) |>
  tab_style(
    style = cell_text(style = "italic", size = px(12)),
    locations = cells_footnotes()
  )
```
