---
author: Nasir Abdulrasheed
title: Data Extraction for OpenPrescribing Seasonality
execute: 
  echo: true
format:
  html: 
    toc: true
    colorlinks: true
    number-sections: true
    fig-pos: H
    smooth-scroll: true
    respect-user-color-scheme: true
    theme:
     light: litera
     dark: cyborg
---

# Setup
Load packages and set up variables for BNF codes and data date range (May 2016 to April 2026).

::: {.callout-warning}
Before running the notebook for data fetching, kindly check the dates and BNF codes to ensure you are not wasting BigQuery resources.
:::

```{r}
#| label: setup
#| warning: false

library(tidyverse)
library(opr)
library(arrow)
library(gt)
```


```{r}
#| label: variables

BNF_ANTIDEPRESSANTS <- "0403%"
BNF_ANTIHISTAMINES <- "030401%"
BNF_NORETHISTERONE <- "0604012P0%"
BNF_ANTIPSYCHOTICS <- "040202%"

DATE_START <- "2016-05-01"
DATE_END <- "2016-06-01"
```

# Drug definition

@tbl-bnf-codes shows the list of chemicals and sections whose data will be extracted for this study.

```{r}
#| label: tbl-product-codes
#| tbl-cap: BNF codes for the drugs/drug groups used in this study
#| code-fold: true

codes <- c(
  "0403*******", "030401*****",
  "0604012P0", "040202*****"
)
drug_names <- c(
  "Antidepressants", "Antihistamines",
  "Norethisterone", "Depot Antipsychotics"
)

data.frame(
  `Chemical/Section Name` = drug_names,
  `BNF Code` = codes,
  check.names = FALSE
) |>
  gt() |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )

```


Connect to BigQuery. See [here](https://bennettoxford.github.io/opr/articles/opr.html#connect).

```{r}
#| label: bq-connect
#| output: false

CONN <- connect_bq()
```

# Fetch data
Query the `normalised_prescribing` table to fetch the data for the BNF codes. The function below takes the connection object and either display the query or collect the data, based on the `display_query` argument:

```{r}
#| label: fetch-functions

fetch_presc_data <- function(
  conn,
  code,
  start,
  end,
  display_query = TRUE
) {
    df <- get_normalised_prescribing(
        con = conn,
        bnf_codes = code,
        start_date = start,
        end_date = end
      ) |>
      select(-net_cost, -actual_cost)
    
    if (display_query) {
      show_query(df)
    } else {
      collect(df)
    }
}

# fetches given products details from the bnf table
fetch_bnf_codes <- function(bnf_codes, display_query = TRUE) {
  df <- tbl(CONN, "bnf") |>
    filter(presentation_code %in% bnf_codes) |>
    select(-ends_with("_code"), presentation_code)

  if (display_query) {
    show_query(df)
  } else {
    collect(df)
  }
}
```

This emits the query for fetching the antidepressants data as an example:
```{r}
#| label: sql-fetch-presc

fetch_presc_data(
  CONN,
  BNF_ANTIDEPRESSANTS,
  DATE_START,
  DATE_END,
  display_query = TRUE
)
```

## Antidepressants
Fetch the data for antidepressants.

```{r}
#| label: antidepressants

df_antidepressants <- fetch_presc_data(
  CONN,
  BNF_ANTIDEPRESSANTS,
  DATE_START,
  DATE_END,
  display_query = FALSE
)
```

```{r}
#| label: tbl-antidepressants
#| code-fold: true

gt(head(df_antidepressants, 5))
```

Writing to a parquet file.
```{r}
#| label: save-antidepressants

write_parquet(
  df_antidepressants,
  paste0("data/antidepressants_", DATE_START, "_",  DATE_END, ".parquet")
)
```

## Antihistamines
```{r}
#| label: antihistamines

df_antihistamines <- fetch_presc_data(
  CONN,
  BNF_ANTIHISTAMINES,
  DATE_START,
  DATE_END,
  display_query = FALSE
)
```

```{r}
#| label: tbl-antihistamines
#| code-fold: true

gt(head(df_antihistamines, 5))
```

```{r}
#| label: save-antihistamines

write_parquet(
  df_antihistamines,
  paste0("data/antihistamines_", DATE_START, "_",  DATE_END, ".parquet")
)
```

## Depot Antipsychotics
```{r}
#| label: antipsychotics

df_antipsychotics <- fetch_presc_data(
  CONN,
  BNF_ANTIPSYCHOTICS,
  DATE_START,
  DATE_END,
  display_query = FALSE
)
```

```{r}
#| label: tbl-antipsychotics
#| code-fold: true

gt(head(df_antipsychotics, 5))
```

```{r}
#| label:  save-antipsychotics

write_parquet(
  df_antipsychotics,
  paste0("data/antipsychotics_", DATE_START, "_",  DATE_END, ".parquet")
)
```

## Norethisterone
```{r}
#| label: norethisterone

df_norethisterone <- fetch_presc_data(
  CONN,
  BNF_NORETHISTERONE,
  DATE_START,
  DATE_END,
  display_query = FALSE
)
```

```{r}
#| label: tbl-norethisterone
#| code-fold: true

gt(head(df_norethisterone, 5))
```

```{r}
#| label: save-norethisterone

write_parquet(
  df_norethisterone,
  paste0("data/norethisterone_", DATE_START, "_",  DATE_END, ".parquet")
)
```

## BNF details

Fetch details (mostly names) of each medicinal product/presentations included in all the data extracted:
```{r}
#| label: sql-fetch-bnf-codes

c(
  df_antidepressants$bnf_code,
  df_antihistamines$bnf_code,
  df_antipsychotics$bnf_code,
  df_norethisterone$bnf_code
) |>
  unique() |>
  fetch_bnf_codes(display_query = TRUE)
```

```{r}
#| label: tbl-bnf-codes
#| tbl-cap: BNF details of drugs in the fetched data
#| code-fold: true

df_bnf_codes <- c(
    df_antidepressants$bnf_code,
    df_antihistamines$bnf_code,
    df_antipsychotics$bnf_code,
    df_norethisterone$bnf_code
  ) |>
    unique() |>
    fetch_bnf_codes(display_query = FALSE)

write_parquet(
  df_bnf_codes,
  paste0("data/bnf_", DATE_START, "_",  DATE_END, ".parquet")
)

head(df_bnf_codes, 5) |> gt()
```

## NHSE regions
Fetch all NHS regions for likely sensitivity analysis later on:

```{r}
#| label: tbl-regions
#| tbl-cap: All NHSE commissioning regions
#| code-fold: true

df_regions <- tbl(CONN, "regional_teams") |>
  collect()

write_parquet(df_regions, paste0("data/regional_teams.parquet"))

gt(df_regions)
```
