---
author: Nasir Abdulrasheed
title: Data Extraction for OpenPrescribing Seasonality
sidebar: nasir
execute: 
  echo: true
  eval: false
format:
  html: 
    toc: true
    colorlinks: true
    number-sections: true
    fig-pos: H
    smooth-scroll: true
    respect-user-color-scheme: true
    theme:
     light: litera
     dark: cyborg
---

# Setup
Load packages and set up variables for BNF codes and data date range (May 2016 to April 2026).

::: {.callout-warning}
Before running the notebook for data fetching, kindly check the dates and BNF codes to ensure you are not wasting BigQuery resources.
:::

```{r}
#| label: setup
#| warning: false
#| code-fold: true

library(tidyverse)
library(opr)
library(arrow)
library(gt)
library(here)
library(glue)
```

```{r}
#| label: variables

# dev set of 4 drugs
SERTRALINE_HCL <- "0403030Q0%"
ESCITALOPRAM <- "0403030X0%"
LORATADINE <- "0304010D0%"
FEXOFENADINE_HCL <- "0304010E0%"
DEV_CODES <- c(SERTRALINE_HCL, ESCITALOPRAM, FEXOFENADINE_HCL, LORATADINE)

# test drug set
FLUPENTIXOL <- "0402020G0%"
PALIPERIDONE <- "0402020AB%"
ZUCLOPENTHIXOL <- "0402020Z0%"
NORETHISTERONE <- "0604012P0%"
CELECOXIB <- "1001010AH%"
TEST_CODES <- c(
  FLUPENTIXOL,
  PALIPERIDONE,
  ZUCLOPENTHIXOL,
  NORETHISTERONE,
  CELECOXIB
)

DATE_START <- "2016-05-01"
DATE_END <- "2026-04-01"
```

# Drug definition

@tbl-dev-drugs and @tbl-test-drugs show the list of chemicals and sections whose data will be extracted for this study.

# Fetch data
Connect to BigQuery. See [here](https://bennettoxford.github.io/opr/articles/opr.html#connect).

```{r}
#| label: bq-connect
#| output: false

CONN <- connect_bq()
```

Query the `normalised_prescribing` table to fetch the data for the BNF codes. The function below takes the connection object and either display the query or collect the data, based on the `display_query` argument:

```{r}
#| label: fetch-functions

fetch_presc_data <- function(code, display_query = TRUE) {
  df <- get_normalised_prescribing(
    con = CONN,
    bnf_codes = code,
    start_date = DATE_START,
    end_date = DATE_END
  ) |>
    select(-net_cost, -actual_cost)

  if (display_query) {
    show_query(df)
  } else {
    collect(df)
  }
}

# fetches given products details from the bnf table
fetch_bnf_codes <- function(bnf_codes, display_query = TRUE) {
  df <- tbl(CONN, "bnf") |>
    filter(presentation_code %in% bnf_codes) |>
    select(
      -ends_with("_code"),
      -presentation,
      chemical_code,
      presentation_code
    )

  if (display_query) {
    show_query(df)
  } else {
    collect(df)
  }
}
```

This emits the query for fetching the prescribing data for the development drugs as an example:

```{r}
#| label: sql-dev-codes

fetch_presc_data(DEV_CODES, display_query = TRUE)
```

## Development drugs
The data for the four drugs/chemicals in @tbl-dev-drugs will be used in developing the pipeline in the first instance, while other drugs will used for testing after implementation is complete.

```{r}
#| label: tbl-dev-drugs
#| tbl-cap: Drugs for initial pipeline development
#| code-fold: true

drug_names <- c(
  "Sertraline",
  "Escitalopram",
  "Loratadine",
  "Fexofenadine"
)

data.frame(
  `Chemical Name` = drug_names,
  `BNF Code` = DEV_CODES,
  check.names = FALSE
) |>
  gt() |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

```{r}
#| label: tbl-presc_dev
#| tbl-cap: Data for Sertraline, Escitalopram, Loratadine, and fexofenadine
#| code-fold: true

df_presc_dev <- fetch_presc_data(DEV_CODES, display_query = FALSE)

head(df_presc_dev, 5) |> gt()
```

```{r}
#| label: tbl-bnf-dev
#| tbl-cap: BNF details for  Sertraline, Escitalopram, Loratadine, and fexofenadine
#| code-fold: true

df_bnf_dev <- unique(df_presc_dev$bnf_code) |>
  fetch_bnf_codes(display_query = FALSE)

head(df_bnf_dev, 5) |> gt()
```

```{r}
#| label: all-dev-data

df_dev_drugs <- df_presc_dev |>
  left_join(df_bnf_dev, c("bnf_code" = "presentation_code"))

save_file <- glue("df_dev_drugs_{DATE_START}_{DATE_END}.parquet")
write_parquet(
  df_dev_drugs,
  here("data-raw", "nasir", save_file)
)

head(df_dev_drugs, 5) |> gt()
```

## Test drugs

These drugs are used to test the output of the pipeline after full implementation.

```{r}
#| label: tbl-test-drugs
#| tbl-cap: Drugs for testing the pipeline
#| code-fold: true

drug_names <- c(
  "Flupentixol",
  "Paliperidone",
  "Zuclopenthixol",
  "Norethisterone",
  "Celecoxib"
)

data.frame(
  `Chemical Name` = drug_names,
  `BNF Code` = TEST_CODES,
  check.names = FALSE
) |>
  gt() |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

```{r}
#| label: all-test-data
#| tbl-cap: Data for test drugs
#| code-fold: true

df_presc_test <- fetch_presc_data(TEST_CODES, display_query = FALSE)

df_bnf_test <- unique(df_presc_test$bnf_code) |>
  fetch_bnf_codes(display_query = FALSE)

df_test_drugs <- df_presc_test |>
  left_join(df_bnf_test, c("bnf_code" = "presentation_code"))

save_file <- glue("df_test_drugs_{DATE_START}_{DATE_END}.parquet")
write_parquet(
  df_test_drugs,
  here("data-raw", "nasir", save_file)
)

head(df_test_drugs, 5) |> gt()
```

## NHSE regions
Fetch all NHS regions for likely sensitivity analysis later on:

```{r}
#| label: tbl-regions
#| tbl-cap: All NHSE commissioning regions
#| code-fold: true

df_regions <- tbl(CONN, "regional_teams") |> collect()


write_parquet(df_regions, here("data-raw", "nasir", "regional_teams.parquet"))

gt(df_regions)
```
