---
title: "Example project"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Background

This study aims to determine whether the sertraline (BNF codes [`0403030Q0`](https://openprescribing.net/chemical/0403030Q0/)) prescription activity for GP practices in England shows seasonality and whether this seasonality differs across geographical regions.

## Setup

The [`opr` package website](https://bennettoxford.github.io/opr/) explains how to install the package and set up database access.
This analysis is written as a Quarto document.
If Quarto is new to you, the [Quarto guide](https://quarto.org/docs/guide/) is a good place to start.

```{r}
#| label: setup
library(opr)
library(tidyverse)
library(scales)
library(gt)
library(here)
library(arrow)
```

## Connect

The OpenPrescribing data lives in [Google BigQuery](https://cloud.google.com/bigquery).
We can connect to it like this:

```{r}
#| label: connect
con <- connect_bq()
```

## Explore

The `get_*()` functions don't download any data straight away.
They build a query that runs in the database, and nothing is downloaded until you call `collect()`.
This means you can preview and refine a query before pulling the results into R.
For example, preview standard GP practices:

```{r}
#| label: explore-practices
get_practices(con, filter_setting = 4, add_setting_labels = TRUE)
```

The `bnf` table shows which BNF codes belong to sertraline:

```{r}
#| label: explore-bnf
tbl(con, "bnf") |>
  filter(str_detect(chemical, "Sertraline")) |>
  distinct(chemical, chemical_code)
```

Add `%` to the end of the chemical code to match all sertraline products, so all the different doses and forms like tablets or oral solutions.

## Extract

Once you are happy with the data, save a copy with `arrow::write_parquet()` (or `write_csv()`) to `data-raw/example/` and commit it to GitHub (if the files aren't too large).
After that you can work from the saved files instead of querying the database every time: set `eval: false` on the extract chunk below and `eval: true` on the load chunk.

```{r}
#| label: extract-data
#| eval: true
#| cache: true
bnf_code_sertraline <- "0403030Q0%"

df_prescribing <- get_normalised_prescribing(
  con,
  bnf_codes = bnf_code_sertraline,
  start_date = "2022-01-01",
  end_date = "2024-12-01"
) |>
  select(month, practice, bnf_code, bnf_name, items, quantity) |>
  collect()

df_practices <- get_practices(con, filter_setting = 4) |>
  collect()

write_parquet(
  df_prescribing,
  here("data-raw", "example", "df_prescribing.parquet")
)

write_parquet(
  df_practices,
  here("data-raw", "example", "df_practices.parquet")
)
```

```{r}
#| label: load-data
#| eval: false
df_prescribing <- read_parquet(
  here("data-raw", "example", "df_prescribing.parquet")
)

df_practices <- read_parquet(
  here("data-raw", "example", "df_practices.parquet")
)
```

## Analyse

First, add up the items prescribed across England for each month:

```{r}
#| label: get-monthly-items
df_monthly_items <- df_prescribing |>
  group_by(month) |>
  summarise(items = sum(items)) |>
  arrange(month)

df_monthly_items
```

### Figure

```{r}
#| label: fig-monthly-items
#| fig-cap: "Monthly count of prescribed sertraline items, 2022-2024."
#| fig-height: 4
#| fig-width: 8
fig_monthly_items <- df_monthly_items |>
  ggplot(aes(x = month, y = items)) +
  geom_line(alpha = 0.4) +
  geom_point(size = .8) +
  scale_x_date(
    labels = label_date("%b\n%Y"),
    date_minor_breaks = "1 month",
    guide = guide_axis(minor.ticks = TRUE)
  ) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Items")

fig_monthly_items

ggsave(
  filename = here("outputs", "example", "figures", "fig_monthly_items.png"),
  fig_monthly_items,
  height = 4,
  width = 8
)
```

We can also plot one line for each BNF code to compare the different sertraline products:

```{r}
#| label: fig-monthly-items-bnf
#| fig-cap: "Monthly count of prescribed sertraline items by BNF code, 2022-2024."
#| fig-height: 6
#| fig-width: 8
fig_monthly_items_bnf <- df_prescribing |>
  group_by(month, bnf_name) |>
  summarise(items = sum(items)) |>
  ggplot(aes(x = month, y = items, colour = bnf_name)) +
  geom_line(alpha = 0.4) +
  geom_point(size = .8) +
  scale_x_date(
    labels = label_date("%b\n%Y"),
    date_minor_breaks = "1 month",
    guide = guide_axis(minor.ticks = TRUE)
  ) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Items", colour = NULL) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7)
  ) +
  guides(colour = guide_legend(ncol = 3))

fig_monthly_items_bnf

ggsave(
  filename = here("outputs", "example", "figures", "fig_monthly_items_bnf.png"),
  fig_monthly_items_bnf,
  height = 6,
  width = 8
)
```

### Table

```{r}
#| label: tbl-annual-items
#| tbl-cap: "Annual count of prescribed sertraline items, 2022-2024."
tab_annual_items <- df_monthly_items |>
  group_by(year = year(month)) |>
  summarise(items = sum(items)) |>
  gt() |>
  cols_label(
    year = "Year",
    items = "Items"
  ) |>
  fmt_number(columns = items, decimals = 0) |>
  tab_header(title = "Annual sertraline prescribing") |>
  tab_source_note(paste("BNF code:", bnf_code_sertraline))

gtsave(
  tab_annual_items,
  here("outputs", "example", "tables", "tab_annual_items.png")
)

tab_annual_items
```
